\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{microtype}
\hypersetup{colorlinks=true, linkcolor=blue!60!black, citecolor=blue!60!black, urlcolor=blue!60!black}

\newcommand{\bpb}{\textsc{bpb}}
\newcommand{\evid}[1]{\footnote{Evidence: \texttt{#1}}}
\newcommand{\todoev}[1]{\textcolor{red!70!black}{[to regenerate: #1]}}

\title{Strictly Dominated: A Post-Mortem of a Spectral--Associative Language
Model, and the In-Block Causality Leak Behind Its $7.2\times$ Headline}
\author{Akhilesh Gogikar\thanks{Reproduction: the full leak harness runs in
$\sim$35\,s on CPU via \texttt{python3
experiments/wht\_causality/confirm\_wht\_unsalvageable.py} in the accompanying
repository: \texttt{https://github.com/Akhilesh-Gogikar/causal-certificate}.}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We report a negative result. Over several months we developed AKI
(\emph{Associative Koopman Imagination}) and SKC (\emph{Spectral Koopman
Capsules}), a family
of attention-replacement language-model blocks combining blockwise
Walsh--Hadamard (WHT) spectral mixing, associative capsule memory, a Koopman-style
latent rollout, and an episodic ``engram'' sidecar. Early experiments showed a
headline result of $7.2\times$ lower bits-per-byte than a matched Transformer
(0.2455 vs.\ 1.7712 \bpb). We show forensically that this result, and every
below-baseline number produced by the family, is explained by an in-block
causality leak: a transform that is causal \emph{between} 64-position blocks but
dense \emph{within} them, so that each position reads inputs from its own
future. A minimal harness demonstrates that (a) perturbing a future in-block
position moves earlier outputs, and (b) because two WHT applications with a
learned diagonal between them form a dyadic (XOR) convolution containing the
permutation $t \mapsto t \oplus 1$, a two-line model trained on i.i.d.\ tokens
drives validation \bpb{} to 2.07 --- far below the information-theoretic floor of
$\log_2 V = 4.0$ --- by literally reading the next token at every even position.
The only strictly causal repair (lower-triangular masking of the block operator)
destroys the fast transform and the entire advantage. With the leak fixed,
matched-pair runs show the architecture was \emph{strictly dominated}: $+0.094$
\bpb{} worse than a parameter/data/seed-matched Transformer at the longest
horizon while running $1.3$--$5.7\times$ slower per step. We dissect the
methodology failures that let a leak survive months of development ---
proxy-FLOPs ``fairness'' contracts, 10-minute micro-gates, multi-knob
attribution chaos, stability-via-damping, and CI gates calibrated to the
contaminated number --- and distill checks that would have caught it in a day.
\end{abstract}

\section{Introduction}
Sub-quadratic replacements for attention are an active research area, and
spectral token mixing is a recurring ingredient. This note documents a
research program (AKI/SKC) whose central empirical support was an artifact of a
causality violation, and whose leak-free variants never beat a matched
Transformer on quality or speed. We publish it because (i) the failure
mechanism --- a \emph{blockwise} fast transform that looks causal at the block
level while leaking within blocks --- is easy to reintroduce in any
FFT/WHT-style mixer for autoregressive models, and passes shape-level and
loss-goes-down sanity checks; (ii) the exploit mechanism is subtle
(spectral-domain gating composes into a future-reading permutation, so the leak
is not just present but \emph{trainable}); and (iii) the process failures are
generic to compute-constrained architecture research.

Contributions:
\begin{enumerate}
  \item A forensic, reproducible demonstration of the in-block WHT causality
  leak and of its exploitability by SGD (Section~\ref{sec:leak});
  \item Matched-pair evidence that the strictly causal architecture is
  dominated on the quality/speed plane (Section~\ref{sec:matched});
  \item A methodology post-mortem with concrete, low-cost gates ---
  including a five-minute i.i.d.-token oracle test that bounds any causal
  model at $\log_2 V$ \bpb{} --- that we argue should precede any headline
  claim for a new sequence mixer (Section~\ref{sec:method});
  \item \textbf{Our central original contribution:} a numeric
  \emph{perturbation-based strict-causality certificate} --- inject
  perturbations at future positions $t' > t$ and require exact numerical
  invariance of all outputs at positions $\le t$ --- packaged as a
  standalone, architecture-agnostic CI tool
  (\texttt{causal-certificate}). Unlike shape-level or mask-inspection
  checks, it certifies the realized forward computation, and it is what
  ultimately detected the leak described here.
\end{enumerate}

\section{The architecture family, briefly}
\label{sec:arch}
First, the names. \textbf{AKI} stands for \emph{Associative Koopman
Imagination}: \emph{associative} because token content is stored and
retrieved via outer-product (Hebbian-style) key--value bindings rather than
pairwise attention; \emph{Koopman} because latent state is advanced by a
learned, approximately linear operator in a lifted feature space, in the
spirit of Koopman operator theory for nonlinear dynamics; and
\emph{imagination} because that operator is rolled forward several steps
\emph{without consuming new tokens}, producing predicted future latents that
condition the current output. \textbf{SKC} stands for \emph{Spectral Koopman
Capsules}: \emph{spectral} because token mixing happens in a Walsh--Hadamard
transform domain rather than by dot-product attention; \emph{capsules}
because the associative memory is organized into a bank of small,
independently gated slots (each a low-rank key--value store) rather than one
flat matrix. \textbf{AKF} (\emph{Associative Koopman Flow}) names the latent
rollout module itself, and \textbf{Engram} --- borrowing the neuroscience
term for a physical memory trace --- is an episodic sidecar that reads
before it writes, storing compressed summaries of past context outside the
main residual stream.

The AKI/SKC family (twelve versions, V0--V12) shares one invariant: replace
dense self-attention with a strict-causal structured state path. The
load-bearing primitive is \texttt{causal\_wht\_blockwise}: reshape the sequence
into blocks of 64 positions and apply a normalized Walsh--Hadamard butterfly
within each block; blocks are combined by a causal decaying scan. Layers use
the pattern \emph{WHT $\to$ elementwise/gated ops in the spectral domain $\to$
WHT back}, alongside an associative capsule bank (SKC), a Koopman-flow latent
rollout (AKF), and a read-before-write episodic memory (Engram). Full version
genealogy and claim boundaries are preserved in the repository
(\texttt{docs/architecture/AKI\_VERSIONS\_DETAILED.md}).

\section{The causality leak}
\label{sec:leak}

\subsection{Mechanism}
The blockwise WHT applies the full $64\times64$ Hadamard butterfly across all
positions of each block. It is causal \emph{between} blocks and dense
\emph{within} them: the output at position $t$ mixes inputs at $t' > t$ up to
the end of $t$'s block. The blockwise framing is precisely why it survived
review: block-level dataflow diagrams look causal, and the GPU path (a Triton
kernel computing the same dense $H x$ per block) inherits the identical leak.

\subsection{Test A: future perturbations move earlier outputs}
Perturbing in-block position $t'{=}40$ changes the output at positions
$t<40$ by up to $0.125$ under the repository implementation, and by exactly
$0.0$ under the lower-triangular-masked variant --- the only strictly causal
repair.\evid{experiments/wht\_causality/report.json (regenerated; torch 2.2.2,
seed-pinned)}

\subsection{Test B: the leak is not just present but trainable}
\label{sec:testb}
Why did the leak convert into (fictitiously) excellent language modeling?
Two WHT applications with a per-position diagonal $s$ between them ---
exactly the layer shape used throughout the family --- compute
$H\,\mathrm{diag}(s)\,H$, a \emph{dyadic (XOR) convolution}. Choosing $s$ to
be the Hadamard spectrum of $\delta_1$ yields $y[t] = x[t \oplus 1]$: at every
even position the mixer outputs the embedding of the \emph{next} token. SGD
finds this optimum readily. On i.i.d.\ uniform tokens with $V{=}16$, where any
causal model is information-theoretically pinned at $\log_2 16 = 4.0$ \bpb:

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
Mixer & val \bpb & vs.\ floor $4.0$\\
\midrule
Repository blockwise WHT (leaky) & $\mathbf{2.0721}$ & impossible without future access\\
Lower-triangular-masked WHT & $4.0021$ & at floor\\
No WHT (diagonal only) & $4.0029$ & at floor\\
\bottomrule
\end{tabular}
\caption{I.i.d.-token oracle test. The leaky mixer reads token $t{+}1$
perfectly at even positions ($\approx 2.0$ \bpb{} $=$ half the positions at 0
bits, half at 4). Deeper stacks of the same primitive reach $\approx 0$
\bpb.\protect\evid{experiments/wht\_causality/confirm\_wht\_unsalvageable.py}}
\label{tab:oracle}
\end{table}

This explains the provisional-patent-era headline (0.2455 vs.\ 1.7712 \bpb,
``$7.2\times$''): the model was grading itself on an open-book exam.

\subsection{No causal-and-fast-and-spectral repair exists}
Masking the future entries of the block operator ($H \mapsto
\mathrm{tril}(H)$) restores causality but destroys the butterfly
factorization, the $O(B\log B)$ cost, and empirically the entire advantage:
clean strictly causal re-runs were $+0.09$ to $+0.24$ \bpb{} \emph{worse} than
baseline.\evid{docs/contribution\_fwht\_causality.md;
docs/architecture/POST\_MORTEM\_WHY\_NON\_SOTA.md}

\section{Matched evaluation of the leak-free architecture}
\label{sec:matched}
With strict causality enforced, paired runs trained the AKI-family architecture
and a Transformer under one harness with matched parameters, optimizer, data,
and seed (\texttt{fairness\_contract.json}):

\begin{table}[h]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrrrr}
\toprule
run & steps & AKI \bpb & xfmr \bpb & $\Delta$\bpb & AKI ms/step & xfmr ms/step & slowdown\\
\midrule
\texttt{aki\_vs\_transformer\_10k} & 10000 & 1.6576 & 1.5635 & $+0.094$ & 170.6 & 89.7 & $1.9\times$\\
\texttt{...125m\_fastpath\_d512} & 4351 & 1.6555 & 1.5004 & $+0.155$ & 494.4 & 392.7 & $1.3\times$\\
\texttt{aki\_l32\_gate\_1500} & 1401 & 2.3277 & 2.0198 & $+0.308$ & 121.6 & 64.6 & $1.9\times$\\
\texttt{aki\_d480\_h8\_speed\_gate} & 1451 & 1.9888 & 2.0188 & $-0.030$ & 207.8 & 64.4 & $3.2\times$\\
\texttt{...aki30pct\_d480\_lowrank} & 1451 & 1.9920 & 2.0208 & $-0.029$ & 368.9 & 64.3 & $5.7\times$\\
\bottomrule
\end{tabular}
\caption{Matched pairs (params/optimizer/data/seed). The only ``wins'' are
$-0.03$ \bpb{} at $\sim$1450 steps --- inside optimization noise and bought at
$3$--$6\times$ step cost. At the longest horizon the deficit is $+0.094$ \bpb{}
against a $\sim2\times$ faster baseline, and the gap trend is stable, not
closing.\protect\evid{docs/architecture/POST\_MORTEM\_WHY\_NON\_SOTA.md;
results/aki\_gap\_trend\_20260510/}}
\label{tab:matched}
\end{table}

Two readings matter. First, \emph{quality}: the deficit at 10k steps is stable
(\texttt{gap\_trend.tsv}), so this is not a transient. Second, \emph{speed}:
the architecture was pitched as sub-quadratic-therefore-faster, yet measured
$1.3$--$5.7\times$ slower per step at practical sequence lengths --- the serial
scan, folded rollout, sidecar memory, and gate machinery cost more wall-clock
than the dense attention they replaced. The architecture was on the wrong side
of the quality/speed frontier in every matched run: strictly dominated.

Root cause, in one sentence: every mechanism in the family mixes the past
through a fixed-width, content-agnostic bottleneck, while attention's
content-based addressing over an unbounded cache is exactly what byte-level
language modeling rewards at these scales.

\section{How a leak survives months: methodology post-mortem}
\label{sec:method}
\begin{enumerate}
  \item \textbf{Proxy-FLOPs fairness was fictional.} The contract used
  $\mathrm{proxy\_flops} = 6 \cdot \mathrm{params} \cdot \mathrm{tokens}$,
  which assumes a Transformer compute profile and is blind to serial per-token
  cost. ``Matched FLOPs'' hid a $1.3$--$5.7\times$ wall-clock disadvantage.
  \emph{Gate on wall-clock on identical hardware.}
  \item \textbf{A mis-calibrated baseline manufactured false hope.} An early
  ``$16\times$ convergence'' headline compared against a Transformer sitting at
  cross-entropy $\approx 5.08$ after 800 steps --- barely below the
  $\ln 256 \approx 5.55$ random floor for bytes, i.e.\ an under-tuned baseline.
  Months of roadmap chased this artifact. \emph{Sanity-check the baseline
  against known-good curves before comparing.}
  \item \textbf{Micro-gate horizons.} Architectures were screened at 10-minute
  /\ 1--2k-step gates where init and LR-schedule transients dominate; the only
  ``wins'' in Table~\ref{tab:matched} live there, and vanish by 10k steps.
  \item \textbf{Attribution chaos.} Runs toggled Engram, speculator, capsules,
  MoE, feedback, kernels, halting, and schedules simultaneously; no run
  isolated the architecture, so no gain was attributable and no loss
  localizable.
  \item \textbf{Stability-via-damping removed the thesis.} Each instability
  was met with a clamp (gate max $0.35$, correction radius $0.30$, control
  scale $0.015$, engine gain $0.20$), converging the novel block toward ``a
  small residual tweak on the backbone'' --- quietly deleting the mechanism
  that was supposed to win.
  \item \textbf{The contaminated number became a CI gate.} The benchmark
  contract enforces \texttt{max\_eval\_bpb: 0.35} --- a threshold only a leaky
  model can meet at this scale --- institutionalizing the artifact as the
  definition of success.\evid{configs/benchmarks/skc\_vs\_mamba\_vs\_transformers.json}
  \item \textbf{The check that catches it in five minutes.} Train the smallest
  model containing the new mixer on i.i.d.\ uniform tokens. Any validation
  \bpb{} below $\log_2 V$ is proof of non-causality (Table~\ref{tab:oracle}).
  We recommend this oracle as a standing pre-benchmark gate for any proposed
  sequence mixer, alongside a perturbation probe (Test A) run on \emph{both}
  the reference and accelerated kernel paths.
\end{enumerate}

\section{Related work}
Non-causal spectral token mixing is well established for \emph{encoders}
(e.g.\ Fourier-mixing architectures); the failure documented here arises
specifically when such mixers are ported to autoregressive decoding behind a
block-causal wrapper. State-space and long-convolution models obtain causal
spectral mixing correctly via causal kernels/scans; our result is not evidence
against those designs, only against unmasked in-block fast transforms. While
ad-hoc probes of causal masking appear in engineering folklore, we are not
aware of prior work formalizing a numeric perturbation-invariance
\emph{certificate} of strict causality as a reusable, model-agnostic CI gate;
we claim the certificate, and the perturbation analysis it packages, as
original contributions of this work and release them at
\texttt{https://github.com/Akhilesh-Gogikar/causal-certificate}. We keep
citations minimal deliberately: an internal audit found earlier agent-generated
prior-art lists unreliable, and we cite only what we have verified.

\section{Limitations and scope}
Results are at small scale (byte-level, $\le$125M-parameter class, sequence
lengths where dense attention is fast); we do not claim the family is dominated
at all scales, only that no evidence supports the original claims. The matched-run
table is reproduced from the repository post-mortem; raw logs for three of the
five rows predate the current branch, and we are regenerating the headline pair
and the Triton-path leak confirmation on commodity cloud GPUs
\todoev{RunPod matched gate + GPU/Triton Test A, this revision}.

\section{Conclusion}
The publishable value of this program is the autopsy: a precise, reproducible
mechanism by which a blockwise fast transform silently reads the future, a
demonstration that spectral gating makes such leaks \emph{trainable} rather
than merely present, matched evidence that the honest architecture lost on both
axes, and a five-minute oracle test that converts this class of error from a
months-scale trap into a pre-flight check.

\paragraph{Reproducibility.} \texttt{experiments/wht\_causality/} regenerates
Tables~\ref{tab:oracle} and Test A on CPU in $\sim$35\,s. Matched-gate
reproduction instructions: \texttt{docs/RUNPOD\_REPRO.md}.

\end{document}
